---
title: "final_data_summary"
author: "Jessica Chung /Melissa Carew"
date: "01/03/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This final section enables data to be arranged into a final table for future analysis.

# producing final data summary

Starting from here will need the thershold list and the sample_df from Cache03
```{r}
load(here::here("results/cache/03.RData"))
#save(thresholds_list, file = here("cache/thresholds_list.RData"))
#saveRDS(samples_df, file = here("cache/samples_df.rds"))

# clear Global environment
#load(here::here("cache/thresholds_list.RData"))
#load(here::here("cache/samples_df.rds"))
```

# load results files from previuos steps
```{r}
short_lulu <- readRDS(here("results/lulu/short_lulu.rds"))
right_lulu <- readRDS(here("results/lulu/right_lulu.rds"))
short_classifications <- read_qza(here("results/classification/short_classification.qza"))
right_classifications <- read_qza(here("results/classification/right_classification.qza"))
short_seq <- read_qza(here("results/dada2/short_representative_sequences.qza"))
right_seq <- read_qza(here("results/dada2/right_representative_sequences.qza"))
```

# check classification data for short amplicon
```{r}
short_classifications$data %>% head(100)
```

# check classification data for right amplicon
```{r}
right_classifications$data %>% head(100)
```


```{r}
# Pull out table of read counts with LULU filter applied
short_lulu$curated_table

#short_lulu <- data.frame(short_lulu$curated_table)
#short_class <- data.frame(short_classifications$data)
```


```{r}
# Pull out table of read counts with LULU filter applied
right_lulu$curated_table

#right_lulu <- data.frame(right_lulu$curated_table)
#right_class <- data.frame(right_classifications$data)
```

```{r}
# merge short_classifications with tHersholds
short_df <- merge(short_classifications$data, thresholds_list[["short"]], by.x="Feature.ID", by.y="feature-id")

right_df <- merge(right_classifications$data, thresholds_list[["right"]], by.x="Feature.ID", by.y="feature-id")
```

```{r}
short_seq$data %>% head
```

```{r}
right_seq$data %>% head
```

```{r}
short_df %>% arrange(desc(max_p_identity)) %>% head(100)
```

```{r}
right_df %>% arrange(desc(max_p_identity)) %>% head(100)
```

```{r}
table(short_df$Taxon == "Unassigned")
```

```{r}
table(right_df$Taxon == "Unassigned")
```

```{r}
short_df %>% filter(Taxon == "Unassigned") %>% pull(max_p_identity) %>% table
```

```{r}
right_df %>% filter(Taxon == "Unassigned") %>% pull(max_p_identity) %>% table
```


```{r}
#short_search_results <- read_qza(here("results/classification/preliminary/short_prelim_search_results.qza"))

# for big dataset
short_search_results <- read_tsv(here("results/classification/preliminary/short_vsearch.tsv"))

```

```{r}
short_search_results$data %>% dim
```

```{r}
# right_search_results <- read_qza(here("results/classification/preliminary/right_prelim_search_results.qza"))

# for big dataset
right_search_results <- read_tsv(here("results/classification/preliminary/right_vsearch.tsv"))

```

```{r}
#right_search_results$data %>% dim
```

```{r}
# ncbi_cois <- read.table(db_coi_ids[["short"]], sep="\t")
# colnames(ncbi_cois) <- c("id", "taxonomy")
```


```{r}
# short_hits <- merge(short_search_results$data, ncbi_cois, by.x="V2", by.y="id", all.x=TRUE)
```

## Combine workflow outputs

# fix sample names

This can be changed with future feedback.  

We could use https://docs.qiime2.org/2023.5/plugins/available/feature-table/filter-features-conditionally/
if we want simple filtering, but if we want to do something complex, better to do it manually.


```{r}
all_tables <- list(
  short=short_lulu$curated_table,
  right=right_lulu$curated_table
)
```


```{r}
# Get sample names and groups
sample_groups <- samples_df$factor
names(sample_groups) <- samples_df$sample_name
sample_groups
```

```{r}
# Need to rename columns to get rid of the prefix X character if sample names start with a digit
#if (all(! colnames(all_tables[[1]]) %in% names(sample_groups))) {
#  all_tables <- lapply(all_tables, function(x) {
#    colnames(x) <- str_remove(colnames(x), "^X")
#   return(x)
#  })
#}

```

# combining read tables for the short and right amplicons
```{r}
#make them dataframes
short_read_tab <- all_tables$short
right_read_tab <- all_tables$right
```

```{r}
# Do the amplicon datasets contain the same number of samples?
short_read_tab %>% dim
right_read_tab %>% dim

# are these samples the same for both amplicons?
if(identical(names(short_read_tab), names(right_read_tab))) {
  print("Column names match.")
} else {
  print("Column names do not match.")
}
```

# if column names match skip this step.
```{r}
# if they do not match which columns are different?
names_only_in_short_read_tab <- setdiff(names(short_read_tab), names(right_read_tab))
print("Column names present only in short_read_tab:")
print(names_only_in_short_read_tab)

names_only_in_right_read_tab <- setdiff(names(right_read_tab), names(short_read_tab))
print("Column names present only in right_read_tab:")
print(names_only_in_right_read_tab)
```

# write scripts to add a blank column to align data for all amplicons
```{r}
# align data set by adding blank column for columns not present in one amplicon but not the other (Cont3rep1, Cont1rep2). 
#short_read_tab <- mutate(short_read_tab, Cont3rep1 = c(0))
#right_read_tab <- mutate(right_read_tab, Cont1rep2 = c(0))
```

```{r}
#merge short and right reads tables into one dataframe
all_read_df <-  rbind(short_read_tab, right_read_tab)

# covert row headers to column and then name column 1
all_read_df <- rownames_to_column(all_read_df, var = "asv_code")
```

#Combining read classification tables for the short and right amplicons
```{r}
# bind short and right classification data
all_classif_df <-  rbind(short_df, right_df)

# rename column 1
colnames(all_classif_df)[1] <- "asv_code"

# join into a singe dataframe
all_readclass_df <- left_join(all_classif_df, all_read_df, by = "asv_code")
```

# add in ASV sequence data
```{r}
# put sequences in a dataframe
short_asvseq <- data.frame(short_seq$data)
right_asvseq <- data.frame(right_seq$data)

# covert row headers to column and then name column 1
short_asvseq <- rownames_to_column(short_asvseq, var = "asv_code")
right_asvseq <- rownames_to_column(right_asvseq, var = "asv_code")

# add amplicon information
short_asvseq  <- mutate(short_asvseq, amplicon = "short")
right_asvseq <- mutate(right_asvseq, amplicon = "right")

# align column names for merging
short_asvseq <- short_asvseq %>% rename(asv_seq = short_seq.data)
right_asvseq <- right_asvseq %>% rename(asv_seq = right_seq.data)


# merge data into a single dataframe
all_asvseq <-  rbind(short_asvseq, right_asvseq)
```

```{r}
# merge ALL data into a single dataframe
final_df <- left_join(all_readclass_df, all_asvseq, by = "asv_code")

# remove columns without data
final_df <- na.omit(final_df)

final_df$Taxon <- gsub("k__", "", final_df$Taxon)
final_df$Taxon <- gsub("p__", "", final_df$Taxon)
final_df$Taxon <- gsub("c__", "", final_df$Taxon)
final_df$Taxon <- gsub("o__", "", final_df$Taxon)
final_df$Taxon <- gsub("f__", "", final_df$Taxon)
final_df$Taxon <- gsub("g__", "", final_df$Taxon)
final_df$Taxon <- gsub("s__", "", final_df$Taxon)

# place taxonomic data in separate columns
final_df <- separate(final_df, Taxon, into = c("kingdom", "phylum", "class", "order", "family", "genus","species"), sep = ";",  fill = "right")

# final tidy up (remove 'NA' and X from start of sample names)
final_df[is.na(final_df)] <- ""
names(final_df) <- sub("X", "", names(final_df))
head(final_df)
```


```{r}
# save final dataframe
write.csv(final_df, file = "~/git/metabarcoding_workflow/results/XXXXX_final_data_summaryX_XXX_24.csv", row.names = FALSE)
```


